Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections

### 2023-07-10 paper.tex
failed attempt to go to springer nature template for Computational Statistics.

\documentclass[
]{jss}

@@

\author{
Nicholas Spyrison~\orcidlink{0000-0002-8417-0212}\\Monash
University \And Dianne Cook~\orcidlink{0000-0002-3813-7155}\\Monash
University \And Przemyslaw
Biecek~\orcidlink{0000-0001-8423-1823}\\University of Warsaw\\
Warsaw University of Technology
}
\title{Exploring Local Explanations of Nonlinear Models Using Animated
Linear Projections}

\Plainauthor{Nicholas Spyrison, Dianne Cook, Przemyslaw Biecek}
\Plaintitle{Exploring Local Explanations of Nonlinear Models Using
Animated Linear Projections}
\Shorttitle{Exploring Local Explanations}

\Abstract{
The increased predictive power of machine learning models comes at the
cost of increased complexity and loss of interpretability, particularly
in comparison to parametric statistical models. This trade-off has led
to the emergence of eXplainable AI (XAI) which provides methods, such as
local explanations (LEs) and local variable attributions (LVAs), to shed
light on how a model use predictors to arrive at a prediction. These
provide a point estimate of the linear variable importance in the
vicinity of a single observation. However, LVAs tend not to effectively
handle association between predictors. To understand how the interaction
between predictors affects the variable importance estimate, we can
convert LVAs into linear projections and use the radial tour. This is
also useful for learning how a model has made a mistake, or the effect
of outliers, or clustering of observations. The approach is llustrated
with examples from categorical (penguin species, chocolate types) and
quantitative (soccer/football salaries, house prices) response models.
The methods are implemented in the R package cheem, available on CRAN.
}

@@

An implementation is provided in the open-source \textbf{R} package
\textbf{cheem}, available on CRAN at
\url{https://CRAN.R-project.org/package=cheem}. Example data sets are
provided, and you can upload your data after model fitting and computing
the LVAs. The LVAs need to be pre-computed and uploaded. Examples show
how to do this for tree SHAP values, using \textbf{treeshap} (tree based
models from \textbf{gbm}, \textbf{lightgbm}, \textbf{randomForest},
\textbf{ranger}, or \textbf{xgboost} \citet{greenwell_gbm_2020};
\citet{shi_lightgbm_2022}; \citet{liaw_classification_2002};
\citet{wright_ranger_2017}; \citet{chen_xgboost_2021}, respectively).
The SHAP and oscillation explanations could be easily added using
\texttt{DALEX::explain()}
\citep{biecek_dalex_2018, biecek_explanatory_2021}.

The application was made with \textbf{shiny} \citep{chang_shiny_2021}.
The tour visual is built with \textbf{spinifex}
\citep{spyrison_spinifex_2020}. Both views are created first with
\textbf{ggplot2} \citep{wickham_ggplot2_2016} and then rendered as
interactive \texttt{html} widgets with \textbf{plotly}
\citep{sievert_interactive_2020}. \textbf{DALEX}
\citep{biecek_dalex_2018} and \emph{Explanatory Model Analysis}
\citep{biecek_explanatory_2021} are helpfule for understanding LEs and
how to apply them.

>>>>

\documentclass[
]{sn-jnl}

@@

\author{
Nicholas Spyrison~\orcidlink{0000-0002-8417-0212}\\Monash
University \And Dianne Cook~\orcidlink{0000-0002-3813-7155}\\Monash
University \And Przemyslaw
Biecek~\orcidlink{0000-0001-8423-1823}\\University of Warsaw\\
Warsaw University of Technology
}
\title{}

\Plainauthor{Nicholas Spyrison, Dianne Cook, Przemyslaw Biecek}

\Abstract{
The increased predictive power of machine learning models comes at the
cost of increased complexity and loss of interpretability, particularly
in comparison to parametric statistical models. This trade-off has led
to the emergence of eXplainable AI (XAI) which provides methods, such as
local explanations (LEs) and local variable attributions (LVAs), to shed
light on how a model use predictors to arrive at a prediction. These
provide a point estimate of the linear variable importance in the
vicinity of a single observation. However, LVAs tend not to effectively
handle association between predictors. To understand how the interaction
between predictors affects the variable importance estimate, we can
convert LVAs into linear projections and use the radial tour. This is
also useful for learning how a model has made a mistake, or the effect
of outliers, or the clustering of observations. The approach is
illustrated with examples from categorical (penguin species, chocolate
types) and quantitative (soccer/football salaries, house prices)
response models. The methods are implemented in the R package cheem,
available on CRAN.
}

@@

An implementation is provided in the open-source \textbf{R} package
\textbf{cheem}, available on CRAN at
\url{https://CRAN.R-project.org/package=cheem}. Example data sets are
provided, and you can upload your data after model fitting and computing
the LVAs. The LVAs need to be pre-computed and uploaded. Examples show
how to do this for tree SHAP values, using \textbf{treeshap} (tree-based
models from \textbf{gbm}, \textbf{lightgbm}, \textbf{randomForest},
\textbf{ranger}, or \textbf{xgboost} \citet{greenwell_gbm_2020};
\citet{shi_lightgbm_2022}; \citet{liaw_classification_2002};
\citet{wright_ranger_2017}; \citet{chen_xgboost_2021}, respectively).
The SHAP and oscillation explanations could be easily added using
\texttt{DALEX::explain()}
\citep{biecek_dalex_2018, biecek_explanatory_2021}.

The application was made with \textbf{shiny} \citep{chang_shiny_2021}.
The tour visual is built with \textbf{spinifex}
\citep{spyrison_spinifex_2020}. Both views are created first with
\textbf{ggplot2} \citep{wickham_ggplot2_2016} and then rendered as
interactive \texttt{html} widgets with \textbf{plotly}
\citep{sievert_interactive_2020}. \textbf{DALEX}
\citep{biecek_dalex_2018} and \emph{Explanatory Model Analysis}
\citep{biecek_explanatory_2021} are helpful for understanding LEs and
how to apply them.
